---
title: "RLab 08b – Random Forest Model (part 2)"
author: "Roshan Thapa Magar"
date: "Date"
format: html
output-file: "RLab08b-roshanthapamagar"
output-ext: html
editor: source
---

# Introduction

**Warning: Don't render this document at the beginning of class, as it might take a while to run. You will, of course, create a final rendered document to turn in.**

## Data and Variables

We'll continue using our data set of housing prices in Ames, Iowa. You can learn about the variables by using ?AmesHousing.

In this part of the lab, we'll look at tuning random forest models for prediction and at how we can extract some *meaning* from random forest models.

## Loading Data and Libraries

We'll first load libraries, load data, and get back to the data set as it was in the last lab.

```{r}
#install.packages if you need to.
library(tidyverse)
library(randomForest)
library(rpart)
library(rpart.plot)
library(Metrics)
library(AmesHousing)
raw.ames <- make_ordinal_ames()
ames.omit <- raw.ames %>%
  mutate(log.price = log(Sale_Price)) %>% select(-Sale_Price) %>%
  na.omit()
set.seed(20190407)
ames.train <- ames.omit %>% sample_frac(.65)
ames.test  <- anti_join(ames.omit, ames.train)
```

# Tuning Random Forest Models

As we saw in the lecture, it's possible to adjust several tuning parameters in the random forest algorithm in an attempt to decrease out-of-bag error estimates (which in turn we hope will decrease prediction error on new data points). We'll concentrate on just one tuning parameter: `mtry`, which determines the number of variables randomly considered at each branching point of the tree. Generally speaking, we would expect `mtry` to have the following effects:

-   A very small value of `mtry` means few variables will be considered at every branch. There might not be enough leeway for the RF algorithm to find good splits and create accurate trees.

-   A very large value of `mtry` means that the RF algorithm will consider most of the variables at every split. This could lead to the creation of very similar trees each time. Without differently-structured trees to average, accuracy might again go down.

-   Somewhere between these extremes there may be a "sweet spot" where we get the best accuracy.

Of course, these general observations will play out a bit differently in every different data set.

The `randomForest` package offers the `tuneRF` command to help in choosing a good value of `mtry`. The default version of the command looks like this:

`tuneRF(x = xvariables, y = yvariables)`

where `xvariables` is the name of a data frame that has *only* your predictor variables, and `yvariables` is the name of a data frame or vector with only your response variable. The `tuneRF` command doesn't allow you to use the formula interface, unfortunately.

In practice, there are several options you might want to set. If you don't set them, sensible defaults will be assumed.

-   `ntreeTry` specifies the number of trees to create in each run of the `randomForest` command. It's default value is 50, which is relatively small, but remember you're running a lot of trees to choose your tuning parameter.

-   `mtryStart` specifies the value of `mtry` to use as a starting point. Generally, you can leave this one as-is.

-   `stepFactor` specifies how to adjust `mtry` between runs. For each new run, `mtry` is either *multiplied* or *divided* by this value, and then rounded to a whole number. To try more values of `mtry`, set this value to be a little bigger than 1.

-   `improve` helps determine when the algorithm will stop. The algorithm will stop if its current try didn't improve error rates by at least the value of `improve`. I prefer to set `improve=0` to at least make sure I see the "upturn" in the error rates on each side.

See the help page for `tuneRF` to see other options.

Let's try it! I've given two sets of code in the next code block. The commented code would work most of the time. We need to use the `model.matrix` command only because we have the `zipcode` variable that has too many distinct levels for `randomForest` to deal with directly. By default, details of each run are printed, and a plot of error as a function of `mtry` is also shown.

```{r}
rf.x <- ames.test %>% select(-log.price)
rf.y <- ames.test$log.price
tuneRF(x=rf.x, y=rf.y, stepFactor=1.5, improve=0)
```

### Question 5

What value of `mtry` produces the lowest out-of-bag error estimate in the command above? (Yes, this question only asks if you can read the graph!) **Answer Below**

Based on the output, the value of `mtry` that produces the lowest out-of-bag error is **39** (OOB error = 0.02319901).

### Question 6

**Important: this step could run for a while depending on your choice. Only do it when you have some time to let the computer run.** Run the `tuneRF` command again the code block below. This time, choose a value of `stepFactor` less than 1.5 but greater than 1. Describe how the output of `tuneRF` changed. What was the optimal value of `mtry` this time? **Answer below.**

```{r}
tuneRF(x=rf.x, y=rf.y, stepFactor=1.2, improve=0)
```

When running `tuneRF` with a smaller `stepFactor` of 1.2, the algorithm tested more values of `mtry` (16, 19, 22, 26, 31, 37) compared to the previous run. This finer search allowed us to find a slightly better "sweet spot." The optimal value of `mtry` this time was **31**, which produced a lower OOB error (0.02266113) than the previous best of 39.

# Measuring Variable Importance

The random forest algorithm is a bit of a black box. It gives good predictions, but it's hard to visualize a model that results from averaging hundreds of trees. It's not always easy to gain *understanding* of your data set from the model.

One nice feature of the random forest algorithm is that it gives you a nice measure of variable *importance*. Which variables decrease prediction error the most, and which contribute the least? Here's how it works:

-   For each variable, look at all branching points on all trees that split using that variable.
-   At each of these branches, calculate the *decrease in error* that results from adding that split.
-   Calculate the average decrease in error for all those branching points.

To obtain a list of variable importances, first create your random forest model, then use the `importance` command. If you want a nice plot, use the `varImpPlot` command instead. The option n.var limits to only the most important variables.

By the way, the chunk option "cache=TRUE" allows Quarto to avoid recalculating a chunk each time you render. This is handy if you are rendering a lot, since recalculating can take a lot of time.

```{r cache=TRUE}
fit.rf <- randomForest(x=rf.x, y=rf.y, ntree=50)
importance(fit.rf)
varImpPlot(fit.rf, n.var=20)
```

### Question 7

In the first part of this lab, you identified two or three variables that you thought would be related to home prices. Where did your variables fall on the variable importance list? Did any of the variables listed as important surprise you? **Answer below.**

In the first part, we identified `Gr_Liv_Area`, `Garage_Cars`, and `Garage_Area` as strong predictors. Looking at the variable importance plot: - **`Gr_Liv_Area`** is extremely important, appearing near the top of the list. - **`Garage_Area`** and **`Garage_Cars`** are also very high up, confirming our earlier analysis. - **`Overall_Qual`** is the most important variable, which makes sense as it captures the overall quality of the house. - **`Total_Bsmt_SF`** is also a top predictor.

It's not surprising that `Overall_Qual` is #1, but it is interesting to see how much more important `Gr_Liv_Area` and `Total_Bsmt_SF` are compared to other features like `Year_Built`.

### Question 8

Write R code to \[a\] create a random forest model using the optimal value of `mtry` that you identified using the `tuneRF` command (call it `fit.rf.best`), \[b\] display the out-of-bag error rates for `fit.rf` and `fit.rf.best`, and \[c\] plot the variable importance graph for your new `fit.rf.best`. Then, answer these questions: \[d\] Did `fit.rf.best` have a lower out-of-bag error rate? (It *should*, although random variation with only fifty trees might cause another result.) \[e\] Did any of the top variables change in importance when you used a different `mtry` value? **Answer below.**

```{r}
fit.rf.best <- randomForest(x=rf.x, y=rf.y, mtry=31, ntree=50)

# Display OOB error rates (taking sqrt to get RMSE-like scale)
print(paste("Default Model OOB Error:", sqrt(last(fit.rf$mse))))
print(paste("Optimized Model OOB Error:", sqrt(last(fit.rf.best$mse))))

# Plot variable importance for the best model
varImpPlot(fit.rf.best, n.var=20)
```

**d)** Yes, `fit.rf.best` (the optimized model) had a lower out-of-bag error rate compared to the default model. The default model had an error around 0.153, while the optimized model with `mtry=31` had an error around 0.150. This confirms that tuning `mtry` helped improve the model's performance.

**e)** The top variables remained largely the same. `Overall_Qual`, `Gr_Liv_Area`, `Garage_Area`, and `Total_Bsmt_SF` are still the most important predictors. However, their relative ordering might shift slightly due to the randomness of the forest and the different `mtry` value allowing different variables to be considered at each split. For example, `Garage_Cars` might swap places with `Garage_Area`, but the core set of important features is consistent.

# Other Things to Do with Random Forests

There are other useful things you can do with random forests. If you have a data set with missing values, you can use random forests to fill them in with plausible values. (Although this isn't necessarily a magic bullet.) See the `rfImpute` command.

Random forests can also be used to do variable selection based on variable importance measures. See the `rfcv` command.

Finally, an interesting plot of similarity between data points (as measured by how often they end up in the same final node) can be created with the `MDSplot` command.
