---
title: "Group Project"
author: "Yuchen Hsiao, Roshan Thapa Magar, Shiori Umekawa"
date: today
format: pdf
output-file: "Bank Marketing Dataset Analysis"
output-ext: "pdf"
editor: source
---

```{r}
library(tidyverse)
```

```{r}
bank<-read.csv("bank-additional/bank-additional/bank-additional-full.csv",sep = ";")
```

```{r}
#shows rows and colums
dim(bank)
#shows first few rows
head(bank)
#shows structure(type of variables)
str(bank)
```

```{r}
#summary all variables
summary(bank)
```

```{r}
#checking missing value
colSums(is.na(bank))
```

```{r}
#view unique values per column
sapply(bank,function(x)length(unique(x)))
```

```{r}
#y= if clients are keep subscribing next term deposit 
table(bank$y)
prop.table(table(bank$y))
```

# Data Cleaning

This section documents the data cleaning process. Most cleaning steps can be performed in Excel (see `Excel_Data_Cleaning_Guide.md`). The R code here focuses only on one-hot encoding, which is complex to do efficiently in Excel.

## Minimal R Cleaning (One-Hot Encoding Only)

The following R code performs only the essential one-hot encoding that cannot be efficiently done in Excel. All other cleaning steps should be performed in Excel using the guide provided.

```{r}
# Load required packages
library(tidyverse)

# Load data
bank <- read.csv("bank-additional/bank-additional/bank-additional-full.csv", sep = ";")
cat("Original dataset:", nrow(bank), "rows,", ncol(bank), "columns\n")

# Step 1: Handle "unknown" values (convert to NA for consistency)
bank_clean <- bank
bank_clean[bank_clean == "unknown"] <- NA
cat("Converted 'unknown' to NA values\n")
```

## One-Hot Encoding

This is the main step that cannot be efficiently done in Excel. We convert categorical variables into binary dummy variables.

```{r}
# One-hot encode categorical variables
categorical_vars <- c("job", "marital", "education", "default", "housing", 
                      "loan", "contact", "month", "day_of_week", "poutcome")

bank_encoded <- bank_clean
for (var in categorical_vars) {
  if (var %in% names(bank_encoded)) {
    # Create dummy variables using model.matrix
    temp_var <- factor(bank_encoded[[var]], exclude = NULL)
    dummy_matrix <- model.matrix(~ temp_var)[, -1, drop = FALSE]
    colnames(dummy_matrix) <- sub("^temp_var", "", colnames(dummy_matrix))
    colnames(dummy_matrix) <- paste0(var, "_", colnames(dummy_matrix))
    bank_encoded <- cbind(bank_encoded, dummy_matrix)
  }
}

# Create binary target variable
bank_encoded$y_binary <- ifelse(bank_encoded$y == "yes", 1, 0)

# Remove original categorical columns (keep y for reference)
bank_encoded <- bank_encoded[, !names(bank_encoded) %in% categorical_vars]

cat("Encoding complete. Final dataset:", nrow(bank_encoded), "rows,", ncol(bank_encoded), "columns\n")
```

## Save Cleaned Dataset

The cleaned dataset is saved for further analysis. Normalization and outlier handling can be performed in Excel if needed.

```{r}
# Save the cleaned dataset
write.csv(bank_encoded, "bank_cleaned.csv", row.names = FALSE)
cat("Cleaned dataset saved as 'bank_cleaned.csv'\n")
cat("Final dataset dimensions:", dim(bank_encoded), "\n")

# Display summary
cat("\nDataset Summary:\n")
cat("- Rows:", nrow(bank_encoded), "\n")
cat("- Columns:", ncol(bank_encoded), "\n")
cat("- Missing values:", sum(is.na(bank_encoded)), "\n")
cat("- Target distribution:\n")
print(table(bank_encoded$y, useNA = "ifany"))
```

# Exploratory Data Analysis (Roshan)

This section explores the cleaned dataset through various visualizations to understand distributions, relationships, and trends in the data.

## Histogram Visualizations

Histograms show the distribution of key numerical variables, helping us understand the shape and spread of the data.

```{r}
# Age distribution
ggplot(bank_encoded, aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Client Ages",
       x = "Age", y = "Frequency")
```

```{r}
# Duration distribution
ggplot(bank_encoded, aes(x = duration)) +
  geom_histogram(bins = 50, fill = "coral", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Call Duration (seconds)",
       x = "Duration (seconds)", y = "Frequency")
```

```{r}
# Campaign distribution
ggplot(bank_encoded, aes(x = campaign)) +
  geom_histogram(bins = 20, fill = "lightgreen", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Number of Contacts During Campaign",
       x = "Number of Contacts", y = "Frequency")
```

## Scatter Plot with Regression Line

Scatter plots with regression lines help visualize relationships between pairs of variables and identify potential correlations. We examine several key relationships.

Age vs Duration: Exploring whether client age relates to call duration.

```{r}
# Scatter plot: Age vs Duration with regression line
ggplot(bank_encoded, aes(x = age, y = duration)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  theme_minimal() +
  labs(title = "Relationship Between Age and Call Duration",
       subtitle = "With Linear Regression Line",
       x = "Age", y = "Duration (seconds)")

# Calculate correlation
cor_age_duration <- cor(bank_encoded$age, bank_encoded$duration, use = "complete.obs")
cat("Correlation between age and duration:", round(cor_age_duration, 3), "\n")
```

Duration vs Campaign: Investigating the relationship between call duration and number of campaign contacts.

```{r}
# Scatter plot: Duration vs Campaign with regression line
ggplot(bank_encoded, aes(x = duration, y = campaign)) +
  geom_point(alpha = 0.3, color = "coral") +
  geom_smooth(method = "lm", color = "darkgreen", se = TRUE) +
  theme_minimal() +
  labs(title = "Relationship Between Call Duration and Number of Campaign Contacts",
       subtitle = "With Linear Regression Line",
       x = "Duration (seconds)", y = "Number of Campaign Contacts")

# Calculate correlation
cor_duration_campaign <- cor(bank_encoded$duration, bank_encoded$campaign, use = "complete.obs")
cat("Correlation between duration and campaign:", round(cor_duration_campaign, 3), "\n")
```

Age vs Previous: Examining if age relates to the number of previous campaign contacts.

```{r}
# Scatter plot: Age vs Previous contacts with regression line
ggplot(bank_encoded, aes(x = age, y = previous)) +
  geom_point(alpha = 0.3, color = "purple") +
  geom_smooth(method = "lm", color = "orange", se = TRUE) +
  theme_minimal() +
  labs(title = "Relationship Between Age and Previous Contacts",
       subtitle = "With Linear Regression Line",
       x = "Age", y = "Number of Previous Contacts")

# Calculate correlation
cor_age_previous <- cor(bank_encoded$age, bank_encoded$previous, use = "complete.obs")
cat("Correlation between age and previous contacts:", round(cor_age_previous, 3), "\n")
```

## Box Plot Visualizations

Box plots compare the distribution of numerical variables across different subscription outcomes (yes/no), helping identify differences between subscribers and non-subscribers.

Comparing age distributions between subscribers and non-subscribers.

```{r}
# Box plot: Age by subscription status
ggplot(bank_encoded, aes(x = y, y = age, fill = y)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Age Distribution by Subscription Status",
       x = "Subscription (y)", y = "Age") +
  scale_fill_manual(values = c("no" = "lightcoral", "yes" = "lightgreen"))
```

Comparing call duration distributions between subscribers and non-subscribers.

```{r}
# Box plot: Duration by subscription status
ggplot(bank_encoded, aes(x = y, y = duration, fill = y)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  labs(title = "Call Duration Distribution by Subscription Status",
       x = "Subscription (y)", y = "Duration (seconds)") +
  scale_fill_manual(values = c("no" = "lightcoral", "yes" = "lightgreen"))
```

## Additional Trend Analysis

Summary statistics by subscription status provide quantitative insights into differences between clients who subscribed and those who did not.

```{r}
# Summary statistics of key numerical features by subscription status
library(dplyr)
trend_summary <- bank_encoded %>%
  group_by(y) %>%
  summarise(
    Mean_Age = mean(age, na.rm = TRUE),
    Mean_Duration = mean(duration, na.rm = TRUE),
    Mean_Campaign = mean(campaign, na.rm = TRUE),
    Mean_Previous = mean(previous, na.rm = TRUE),
    .groups = "drop"
  )
print(trend_summary)
```

```{r}
#Distribution of age to see the rang of people who aspect this research 
ggplot(bank, aes(x=age))+
  geom_histogram(fill="steelblue",bins=30,color="pink")+
  theme_minimal()+
  labs(title="Age Distribution of Clients")
```

```{r}
#subscription rate by job type
ggplot(bank, aes(x=job, fill = y))+
  geom_bar(position = "fill")+
  coord_flip()+
  labs(title="Term Desposit Subscription by Job Type", y= "Proportion")
theme_minimal()
```
